Supplementary material to `Adaptive group-regularized logistic elastic net regression'

### 2019-11-29 docs/biostatistics_main.Rnw
added extra simulations

% set options
<<settings, include=FALSE, echo=FALSE>>=
opts_knit$set(root.dir="..", base.dir="../figs")
opts_chunk$set(fig.align='center', fig.path="../figs/", 
               echo=FALSE, cache=FALSE, message=FALSE, fig.pos='!ht')
@

@@

\documentclass[oupdraft]{bio}
\usepackage{xr-hyper}
% \usepackage[colorlinks=true, urlcolor=citecolor, linkcolor=citecolor, 
% citecolor=citecolor]{hyperref}
\usepackage{url,bm,algorithm,algpseudocode,mathtools,bbm,pgfplotstable,
enumerate}

\externaldocument[sm-]{supplement}

@@

  \section{External information sources}\label{sec:external}
  We describe possible sources of external information in omics
  studies that may provide the feature groups $\G(g)$. Firstly, there are 
  biologically motivated partitioning of features that are
  easily retrieved from online repositories. These are already 
  discretised and may be included in the analysis as-is. Examples are:
  (i) pathway memberships of genes,
  (ii) classes of metabolites (SM Section 12),
  and (iii) conservation status of microRNAs (SM Section 
  13).
  A second type of external information comes in the form of continuous data.
  Examples are:
  (iv) $p$-values or false discovery rates (FDRs) from a 
  different, related study (Section \ref{sec:colorectal}), and
  (v) quality scores of the features (SM Section 12), 
  and (vii) the node degrees of a network estimated on the feature 
  data (SM Section 12).

  \texttt{gren} requires discretised external information, so in the case of 
  continuous external data, some form of discretisation is required. In some
  cases discretisation comes naturally. For example, with external data 
  type (iv) one might consider the `standard' cut-offs 0.05, 0.01,
  and 0.001. If the choice of cut-offs is not straightforward, we propose a 
  a data-driven heuristic, which renders a relatively finer discretisation grid 
  for data dense than for data sparse domains. Then,
  adaptation to external information is more pronounced in high density areas.
  We propose to fit a piecewise linear spline to the empirical cumulative
  distribution function of the external data and automatically select knot
  locations as cut-offs \cite[]{spiriti_knot_2013}. They also provide a 
  data-driven method to choose the
  number of knots, for a given maximum number of knots. The maximum number
  of knots should be chosen such that each group contains enough features for 
  stable estimation of the penalty weights. As a rule of thumb, we advise
  at least 20 features per group.

>>>>

% set options
<<settings, include=FALSE, echo=FALSE>>=
opts_knit$set(root.dir="..", base.dir="../figs")
opts_chunk$set(fig.align='center', fig.path="../figs/", 
               echo=FALSE, cache=FALSE, message=FALSE, fig.pos='!ht',
               dev="postscript")
@

@@

\documentclass[oupdraft]{bio}
% \usepackage[colorlinks=true, urlcolor=citecolor, linkcolor=citecolor, 
% citecolor=citecolor]{hyperref}
\usepackage{url,bm,algorithm,algpseudocode,mathtools,bbm,pgfplotstable,
enumerate}

% \externaldocument[sm-]{supplement}

@@

  \section{External information sources}\label{sec:external}
  We describe possible sources of external information in omics
  studies that may provide the feature groups $\G(g)$. Firstly, there are 
  biologically motivated partitioning of features that are
  easily retrieved from online repositories. These are already 
  discretised and may be included in the analysis as-is. Examples are:
  (i) pathway memberships of genes,
  (ii) classes of metabolites (see SM Section 12),
  and (iii) conservation status of microRNAs (see SM Section 
  13).
  A second type of external information comes in the form of continuous data.
  Examples are:
  (iv) $p$-values or false discovery rates (FDRs) from a 
  different, related study (see Section \ref{sec:colorectal}), and
  (v) quality scores of the features (see SM Section 12), 
  and (vii) the node degrees of a network estimated on the feature 
  data (see SM Section 12).

  \texttt{gren} requires discretised external information, so in the case of 
  continuous external data, some form of discretisation is required. In some
  cases discretisation comes naturally. For example, with external data 
  type (iv) one might consider the `standard' cut-offs 0.05, 0.01,
  and 0.001. If the choice of cut-offs is not straightforward, we propose a 
  a data-driven heuristic, which renders a relatively finer discretisation grid 
  for data-dense than for data-sparse domains. Then,
  adaptation to external information is more pronounced in high density areas.
  We propose to fit a piecewise linear spline to the empirical cumulative
  distribution function of the external data and automatically select knot
  locations as cut-offs \cite[]{spiriti_knot_2013}. They also provide a 
  data-driven method to choose the
  number of knots, for a given maximum number of knots. The maximum number
  of knots should be chosen such that each group contains enough features for 
  stable estimation of the penalty weights. As a rule of thumb, we advise
  at least 20 features per group.

### 2019-11-29 docs/manuscript.Rnw
added extra simulations

  The paper is structured as follows. We introduce the model in Section 
  \ref{sec:model}. In Section 
  \ref{sec:external} we shortly discuss possible sources of co-data.
  In Section \ref{sec:estimation} we derive a VB approximation to the
  model introduced in Section \ref{sec:model}
  and use this novel approximation in the empirical Bayes (EB) estimation of 
  multiple, group-specific penalty parameters. In Section \ref{sec:simandapp}
  we compare the method in a simulation study, demonstrate the benefit of the 
  approach for two 
  data sets, and summarise results for two additional data sets. 
  We conclude with a 
  discussion.

@@

	Assume we have a partitioning of the features in $G$ groups, such that each
	feature belongs to one group. Let $\G(g)$ be the feature index set of group 
	$g$ for $g=1, \dots, G$ and let $\lambda'_g \in \mathbb{R}_{>0}$ denote a 
	group-specific penalty weight. In a generalised elastic net regression, the
	penalised likelihood is maximised to yield parameter estimates:
	\begin{equation}\label{eq:weightedlik}
	  \hat{\bbeta} := \underset{\bbeta}{\argmax} \log \L(\y ; \bbeta) -  
	  \frac{\lambda_1}{2} 
	  \sum_{g=1}^G \sqrt{\lambda'_g} \sum_{j \in \G(g)} | \beta_j| - 
	  \frac{\lambda_2}{2} \sum_{g=1}^G \lambda'_g \sum_{j \in \G(g)} \beta_j^2, 
	\end{equation}
	where $\L(\y ; \bbeta)$ denotes the likelihood function of the 
	observed data $\y = \begin{bmatrix} 
	y_1 & \cdots & y_n \end{bmatrix} \tr$, and 
	$\lambda_1, \lambda_2 \in \mathbb{R}_{>0}$ are the `global' penalty 
	parameters. From (\ref{eq:weightedlik}) we see that the
	$\lambda'_g$'s may be interpreted as penalty multipliers. Note that
	the regular elastic net is recovered by setting $\forall g: \lambda'_g=1$.

  Throughout the following we assume that the geometric mean of the multipliers,
  weighted by their 
	respective group sizes, is one, such that the average shrinkage of the model
	parameters is determined by $\lambda_1$ and $\lambda_2$. That is,
	we calibrate the $\lambda'_g$ such that $\prod_{g=1}^G 
	(\lambda'_g)^{|\G(g)|}=1$. The multipliers appear in square root form in the
	$L_1$-norm term to ensure that penalisation on the parameter level scales
	with the norm. The $L_1$-norm sets some of the estimates exactly to zero, thus 
	automatically selecting features. The $L_2$-norm ensures collinearity is 
	well-handled.

@@

  \texttt{gren} requires discretised external information, so in the case of 
  continuous external data, some form of discretisation is required. In some
  cases discretisation comes naturally. For example, with external data 
  type (iv) one might consider the `standard' cut-offs 0.05, 0.01,
  and 0.001. If the choice of cut-offs is not straightforward, we propose a 
  a data-driven heuristic, which renders a relatively finer discretisation grid 
  for data dense than for data sparse domains. Then,
  adaptation to external information is more pronounced in high density areas.
  We propose to fit a piecewise linear spline to the empirical cumulative
  distribution function of the external data and automatically select knot
  locations as cut-offs \cite[]{spiriti_knot_2013}. They also provide a 
  data-driven method to choose the
  number of knots, for a given maximum number of knots. The maximum number
  of knots should be chosen such that each group contains enough features for 
  stable estimation of the penalty weights. As a rule of thumb, we advise
  at least 20 features per group.

@@

>>>>

  The paper is structured as follows. We introduce the model in Section 
  \ref{sec:model}. In Section 
  \ref{sec:external} we shortly discuss possible sources of co-data.
  In Section \ref{sec:estimation} we derive a VB approximation to the
  model introduced in Section \ref{sec:model}
  and use this novel approximation in the empirical Bayes (EB) estimation of 
  multiple, group-specific penalty parameters. In Section \ref{sec:simandapp}
  we compare the method in a simulation study, demonstrate the benefit of the 
  approach for two 
  data sets, and summarise results for two additional data sets. 
  We conclude with a discussion of some of the benefits and drawbacks of the
  proposed \texttt{gren}.

@@

	Assume we have a partitioning of the features in $G$ groups, such that each
	feature belongs to one group. Let $\G(g)$ be the feature index set of group 
	$g$ for $g=1, \dots, G$ and let $\lambda'_g \in \mathbb{R}_{>0}$ denote a 
	group-specific penalty weight. In a generalised elastic net regression, the
	penalised likelihood is maximised to yield parameter estimates:
	\begin{equation}\label{eq:weightedlik}
	  \hat{\bbeta} := \underset{\bbeta}{\argmax} \log \L(\y ; \bbeta) -  
	  \frac{\lambda_1}{2} 
	  \sum_{g=1}^G \sqrt{\lambda'_g} \sum_{j \in \G(g)} | \beta_j| - 
	  \frac{\lambda_2}{2} \sum_{g=1}^G \lambda'_g \sum_{j \in \G(g)} \beta_j^2, 
	\end{equation}
	where $\L(\y ; \bbeta)$ denotes the likelihood function of the 
	observed data $\y = \begin{bmatrix} 
	y_1 & \cdots & y_n \end{bmatrix} \tr$, and 
	$\lambda_1 > 0$ and $\lambda_2 > 0$ are the `global' penalty 
	parameters. From (\ref{eq:weightedlik}) we see that the
	$\lambda'_g$'s may be interpreted as penalty multipliers. Note that
	the regular elastic net is recovered by setting $\forall g: \lambda'_g=1$.

  Throughout the following we assume that the geometric mean of the multipliers,
  weighted by their 
	respective group sizes, is one, such that the average shrinkage of the model
	parameters is determined by $\lambda_1$ and $\lambda_2$. That is,
	we calibrate the $\lambda'_g$ such that $\prod_{g=1}^G 
	(\lambda'_g)^{|\G(g)|}=1$, with $|\G(g)|$ the number of features in group $g$. 
	The multipliers appear in square root form in the
	$L_1$-norm term to ensure that penalisation on the parameter level scales
	with the norm. The $L_1$-norm sets some of the estimates exactly to zero, thus 
	automatically selecting features. The $L_2$-norm ensures collinearity is 
	well-handled. Addition of the penalty terms also prevents quasi-complete
	separation in logistic regression, a common phenomenon in small $n$ studies.

@@

  \texttt{gren} requires discretised external information, so in the case of 
  continuous external data, some form of discretisation is required. In some
  cases discretisation comes naturally. For example, with external data 
  type (iv) one might consider the `standard' cut-offs 0.05, 0.01,
  and 0.001. If the choice of cut-offs is not straightforward, we propose a 
  a data-driven heuristic, which renders a relatively finer discretisation grid 
  for data-dense than for data-sparse domains. Then,
  adaptation to external information is more pronounced in high density areas.
  We propose to fit a piecewise linear spline to the empirical cumulative
  distribution function of the external data and automatically select knot
  locations as cut-offs \cite[]{spiriti_knot_2013}. They also provide a 
  data-driven method to choose the
  number of knots, for a given maximum number of knots. The maximum number
  of knots should be chosen such that each group contains enough features for 
  stable estimation of the penalty weights. As a rule of thumb, we advise
  at least 20 features per group.

@@

	In addition to simulations from the correct model, i.e., model

### 2019-11-29 docs/supplement.Rnw
added extra simulations

>>>>

  \subsection{Model misspecification}
  To assess the effects of model misspecification we conducted a series of 
  simulations under different modes of misspecification and simulation settings.
  We investigate model misspecification by comparing \texttt{gren} to the
  regular elastic net.
  To investigate the effect of misspecifying the link function, we simulated
  according to Scenarios 4 and 5, where we changed the response simulation:
	$y_i \sim \mathcal{B}(1, [\exp (\x_i \tr \bbeta) + \kappa]/
	[\exp (\x_i \tr \bbeta) + 1 + \kappa])$ as in \cite{komori_asymmetric_2016}, 
	where the link function is unbalanced for $\kappa \neq 0$. We use 
	$\kappa \in \{ 0.1, 0.5, 1 \}$, which leads to 6 different simulations 
	from Scenarios 4 and 5 in total. We present the results in Figures
	\ref{fig:lines2_simulations_res6_auc_briers}-
	\ref{fig:lines2_simulations_res11_auc_briers}. From the Figures we see that 
	\texttt{gren} outperforms the regular elastic net even with a misspecified 
	link function in terms of AUC. For the Brier skill score there seems to be 
	some detrimental effect of the misspecification.

<<lines2_simulations_res6_auc_briers, fig.cap="Estimated (a) AUC and (b) Brier skill score in simulation setting 4 for a misspecified link function with $\\kappa=0.1$.", out.width="100%", fig.asp=1/2>>=
@

<<lines2_simulations_res7_auc_briers, fig.cap="Estimated (a) AUC and (b) Brier skill score in simulation setting 4 for a misspecified link function with $\\kappa=0.5$.", out.width="100%", fig.asp=1/2>>=
@

<<lines2_simulations_res8_auc_briers, fig.cap="Estimated (a) AUC and (b) Brier skill score in simulation setting 4 for a misspecified link function with $\\kappa=1$.", out.width="100%", fig.asp=1/2>>=
@

<<lines2_simulations_res9_auc_briers, fig.cap="Estimated (a) AUC and (b) Brier skill score in simulation setting 5 for a misspecified link function with $\\kappa=0.1$.", out.width="100%", fig.asp=1/2>>=
@

<<lines2_simulations_res10_auc_briers, fig.cap="Estimated (a) AUC and (b) Brier skill score in simulation setting 5 for a misspecified link function with $\\kappa=0.5$.", out.width="100%", fig.asp=1/2>>=
@

<<lines2_simulations_res11_auc_briers, fig.cap="Estimated (a) AUC and (b) Brier skill score in simulation setting 5 for a misspecified link function with $\\kappa=1$.", out.width="100%", fig.asp=1/2>>=
@

  In addition to misspecification of the response relation, model 
  misspecification may also occur in the feature relation. To investigate this
  we simulated data that includes non-linear relations between linear predictor
  and features. To do so, we considered the following link function for part of 
  the features: $\expit [\max(0,x_{ij})\beta_j]$. We apply this non-linear 
  relation in simulation Scenarios 4 and 5, where we consider three fractions
  of features for which we apply the non-linear relation: 0.2, 0.4, and 0.6, 
  evenly divided over the groups of features. 
  Meaning that 0.8, 0.6, and 0.4 of the features are linearly related to the
  link function, respectively. This leads again to 6 different simulation 
  settings: three in Scenario 4 and three in Scenario 5. We present the results
  in Figures \ref{fig:lines2_simulations_res12_auc_briers}-
  \ref{fig:lines2_simulations_res14_auc_briers}.
  From the Figures we see that \texttt{gren} still outperforms the regular 
  elastic net, albeit by a slightly smaller margin then with a correctly
  specified feature relation.

<<lines2_simulations_res12_auc_briers, fig.cap="Estimated (a) AUC and (b) Brier skill score in simulation setting 4 for a non-linear feature relation with the fraction of non-linear relations 0.2.", out.width="100%", fig.asp=1/2>>=
@

<<lines2_simulations_res13_auc_briers, fig.cap="Estimated (a) AUC and (b) Brier skill score in simulation setting 4 for a non-linear feature relation with the fraction of non-linear relations 0.4.", out.width="100%", fig.asp=1/2>>=
@

<<lines2_simulations_res14_auc_briers, fig.cap="Estimated (a) AUC and (b) Brier skill score in simulation setting 4 for a non-linear feature relation with the fraction of non-linear relations 0.6.", out.width="100%", fig.asp=1/2>>=
@
